---
title: "R Notebook"
output: html_notebook
editor_options: 
  chunk_output_type: console
---
Analyze the fluorescence intensity of cytosolic Calcium indicator data from Moe-Lange, 2021.

0. Load libraries and functions
```{r}
library(ggplot2)
library(dplyr)
library(tidyr)
# Functions
normalizeToBaseline <- function(trace, bl.time = 90){
  bl.trace = trace$Y[1:which(abs(trace$X-bl.time) ==  min(abs(trace$X-bl.time ))) ]
  bl = round(mean(bl.trace,2))
  trace$Y = trace$Y - bl
  return(trace)
  ## Get slope for each trace. 
}
calculateslope <- function(vector) {
  slopetrace = NULL
  for( i in 2:length(vector)){
    slopei = vector[i]-vector[i-1]
    slopetrace = append(slopetrace, slopei)  
  }
  slopetrace = append(slopetrace, slopetrace[length(slopetrace)])
  return(slopetrace)
}
```


1. Load all the .csv files of the pixel intensity of ROIs in the stacks.

1.1 Setup some data of the experiment.
```{r}
fecha = 210325  ## Date
planta = "Col0" ## line Col0..."Col0-MZ" "Col0-RH-MZ" "Col0-RH-EZ"
individual = 1 ## 
reporter = "MCaMP6s" ## GCaMP3 or "Atto514", "MCaMP6s"
AA =  "none"  
dx = 15 # distance between ROIs in mm
```

1.2 Load and plot
input- traces in .csv format with name indicating basipetal (base) or acropetal (acro). 
By default traces are labaled as L13, if necessary to state a different leaf do it like: "...-base-L13.csv"
```{r}
home = getwd()
list.files()
Alltraces =c() 

for (i in list.files()[grep(".csv", list.files())] ){
    # load data
  trace.i <- read.csv(i)
  if(length( grep("base",i) ) == 1 ){
    trace.i$position = "basipetal"
  } else {
    trace.i$position = "acropetal"
  }
  
  trace.i$name = strsplit(i,"-")[[1]][1]
  trace.i$date = fecha
  trace.i$plant = planta
  trace.i$reporter = reporter
  trace.i$leaf = "13"
  names(trace.i)[1] = "X"
  names(trace.i)[2] = "Y"
  if( length(grep("L",i)) != 0 ){
    trace.i$leaf = strsplit(i,"-")[[1]][3] %>% substr(2,3)
  }
  trace.i <- normalizeToBaseline(trace.i)
  
  ### plot traces  ###
  title = paste( trace.i[1,3:ncol(trace.i)], sep = " ", collapse = " " ) 
  plot(x = trace.i$X, y = trace.i$Y, type = "l", main=title)
  #readline (prompt="Press [enter] to save")
  ### add to file with all data
  Alltraces = rbind(Alltraces, trace.i)
}
# head(Alltraces)
# tail(Alltraces)

write.csv(Alltraces, file = "Alltraces.csv", row.names = FALSE )    
```

2. calculate, plot and save slope traces.
input - Alltraces.csv
output - datasummary.csv
```{r}
# filter for one only trace
datasummary = c()
slope = NULL
for( n in unique(Alltraces$name) ){
  for( p in unique(Alltraces$position)){
    trace.i = dplyr::filter(Alltraces, name==n, position==p) 
    for(l in unique(trace.i$leaf) ){
      trace.i = dplyr::filter(Alltraces, name==n, position==p, leaf==l) 
      print(paste(n,p,l))
      
      # determine baseline
      #plot(trace.i$Y, type="l") + abline(v=100)
      baseline = mean(trace.i$Y[1:100])
      # find slope
      slope.i = calculateslope( trace.i$Y )
      slope = append(slope, slope.i)
      peak.slope = max(slope.i) # save
      
      # find time of Peak slope index and time
      peak.slope.index = which(slope.i==max(slope.i)) 
      peak.slope.time = trace.i$X[peak.slope.index]# save
      
      # find peak amplitude
      peak.amplitude = max(trace.i$Y[1:250])-baseline # save
      peak.amplitude.index = which(trace.i$Y == peak.amplitude+baseline ) 
      peak.amplitude.time = trace.i$X[peak.amplitude.index]# save
      
      # find half max (+ frame)
      halfmax.amplitude = round( peak.amplitude/2 ,2)
      halfmax.frame = which( abs( trace.i$Y[1:(peak.amplitude.index)] - halfmax.amplitude ) == min( abs( trace.i$Y[1:(peak.amplitude.index)] - halfmax.amplitude ) ) ) 
      halfmax.time = trace.i$X[halfmax.frame] # save
      
      # find half max of repol (frame)
      repol.halfmax.frame = which( abs( trace.i$Y[(peak.amplitude.index+5):nrow(trace.i)] - halfmax.amplitude ) == min( abs( trace.i$Y[(peak.amplitude.index+5):nrow(trace.i)] - halfmax.amplitude ) ) ) + peak.amplitude.index+5
      repol.halfmax.time = trace.i$X[repol.halfmax.frame] # save
      

      #plot( trace.i$Y ) + abline(v=repol.halfmax.frame)
      
      # Calculate and save duration
      duration = repol.halfmax.time - halfmax.time # save
      
      
      
      
      # plot
      title = paste( trace.i[1,3:ncol(trace.i)], sep = " ", collapse = " " ) 
      plot(slope.i, type = "l", main=title )+
        abline(v = peak.slope.index, col = "red")
      
      plot(trace.i$X, trace.i$Y, main=title )+
        abline(v = peak.slope.time, col = "red")
      datasummary = rbind(datasummary, data.frame( n,
                                                   p,
                                                   l,
                                                   "peak.slope"=peak.slope,
                                                   "peak.slope.time"=peak.slope.time,
                                                   "peak.amplitude"=peak.amplitude,
                                                   "peak.amplitude.time"=peak.amplitude.time,
                                                   "halfmax.time"=halfmax.time,
                                                   "repol.halfmax.time"=repol.halfmax.time,
                                                   "duration"=duration,
                                                   dx) )
    }
  }
} 
if ( length(grep("slope", names(Alltraces))) == 0){
  Alltraces$Yslope1 = slope
}
write.csv(datasummary, "datasummary.csv", row.names = FALSE )
write.csv(Alltraces, file = "Alltraces.csv", row.names = FALSE )
```
2.1 Make boxplots with the data calculated in 2. 
```{r}
parameters = c("peak.slope", "peak.slope.time", "peak.amplitude", "peak.amplitude.time", "halfmax.time", "repol.halfmax.time", "duration")
for( r in which(names(datasummary)%in%parameters) ){
  ggplot(datasummary,aes(p, datasummary[,r])) +
  geom_boxplot() +
  geom_jitter() +
  ggtitle( names(datasummary[r]) )
  ggsave(paste("Boxplot-",names(datasummary[r]),".pdf",sep = ""))
}

```



3. Calculate and plot speeds of the Ca wave. Use peak slope for this.
OPTIONAL! give a different distance to each experiment. Ideally it's always the same!
```{r}
datasummary$dx = 0
datasummary$dx[which(datasummary$name==4)] = 11
datasummary$dx[which(datasummary$name==6)] = 12
datasummary$dx[which(datasummary$name==5)] = 7
```

input - datasummary.csv
output - plots and datastats.csv
```{r}
speed.data = NULL
for ( n in unique(datasummary$name) ){
  summary.i = dplyr::filter(datasummary, name==n )
  for ( l in unique(summary.i$leaf)) {
    print(paste(n,l,sep = "-"))
    summary.i = dplyr::filter(datasummary, name==n, leaf==l )
    trace.i = dplyr::filter(Alltraces, name==n, leaf==l )
    # find time difference
    dt = summary.i$peak.time[ which(summary.i$position == "acropetal" ) ] - 
        summary.i$peak.time[ which(summary.i$position == "basipetal" ) ]
    # find speed
    speed = round( summary.i$dx[1]/dt , 2)  # mm/s
    # save
    speed.data = rbind( speed.data, 
                        data.frame("name" = n,
                                   "leaf" = l,
                                   "mm/s" = speed) )
    # plot
    plot <- ggplot( trace.i, aes(X, Y, color = position)) +
      geom_line()  +
      geom_vline( xintercept  =summary.i$peak.time[ which(summary.i$position == "acropetal" ) ] ) +
      geom_vline( xintercept = summary.i$peak.time[ which(summary.i$position == "basipetal" ) ]) +
      ylab ("Pixel Intensity ") + 
      xlab ("Seconds") + 
      ggtitle(paste(n,l))
    print(plot)
    ggsave(filename =  paste( n," L",l,"_slopePeaks.pdf",sep = "" ), plot)
  }
}
write.csv(speed.data, "speed.data.csv")

ggplot(speed.data, aes("Col-0",mm.s)) +
  geom_boxplot() + 
  geom_jitter()
ggsave( paste( "speeds.pdf",sep = "" ))
```

4. Curate data manually based on visual inspection of the plots from 3.
```{r}
remove = c("470_1","471_1")

traces.filtered = Alltraces[-c(which(Alltraces$name %in% remove)),]
datasummary = datasummary[-c(which(datasummary$name %in% remove)),]
speed.data = speed.data[-c(which(speed.data$name %in% remove)),]

write.csv(traces.filtered, "traces.filtered.csv", row.names = FALSE)
write.csv(datasummary, "datasummary.csv", row.names = FALSE)
write.csv(speed.data, "speed.data.csv", row.names = FALSE)

ggplot(speed.data, aes("Col-0",mm.s)) +
  geom_boxplot() + 
  geom_jitter()
ggsave( paste( "speeds.pdf",sep = "" ))

```

5. Generate mean ± SEM plot of traces
input - traces.filtered
```{r}
head(Alltraces)
# put all traces in columns
time = dplyr::filter(traces.filtered, name == n & position == "basipetal" ) 
time = time$X[1:301]
basipetal = data.frame("s" = time)
acropetal = data.frame("s" = time)
for (n in unique(traces.filtered$name) ){
  print(n)
  plant.n = dplyr::filter(traces.filtered, name == n)
  # align everything to the peak slope of the basipetal traces.
  normalize.slope = plant.n %>% dplyr::filter(position == "basipetal") %>% summarize(max(Yslope1))
  normalize.frame = which(plant.n$Yslope1 == normalize.slope[[1]]) 
  normalize.time <- plant.n$X[  normalize.frame ]
  plant.n$X <- (plant.n$X - normalize.time)
  ## Pick 300 frames of the Y vector and stack
  basipetal.n = dplyr::filter(plant.n, position == "basipetal")
  basipetal.n = basipetal.n$Y[ (which(basipetal.n$X==0)-50):(which(basipetal.n$X==0)+250) ]
  #
  acropetal.n = dplyr::filter(plant.n, position == "acropetal")
  acropetal.n = acropetal.n$Y[ (which(acropetal.n$X==0)-50):(which(acropetal.n$X==0)+250) ]
  # plot(acropetal.n$Y)
  ## save to collection variables
  basipetal = cbind( basipetal, basipetal.n )
  names(basipetal)[ncol(basipetal)] = n
  acropetal = cbind( acropetal, acropetal.n )
  names(acropetal)[ncol(acropetal)] = n
}
#
basipetal$mean = rowMeans(basipetal[,2:ncol(basipetal)], na.rm = TRUE)
basipetal$pSEM = basipetal$mean + 
      ( apply( basipetal[,2:(ncol(basipetal)-1)], 1, sd, na.rm=TRUE) /
          rowSums(!is.na(basipetal[,2:(ncol(basipetal)-1)]))  )
basipetal$nSEM = basipetal$mean - 
      ( apply( basipetal[,2:(ncol(basipetal)-2)], 1, sd, na.rm=TRUE) /
          rowSums(!is.na(basipetal[,2:(ncol(basipetal)-2)]))  )
#
acropetal$mean = rowMeans(acropetal[,2:ncol(acropetal)], na.rm = TRUE)
acropetal$pSEM = acropetal$mean + 
      ( apply( acropetal[,2:(ncol(acropetal)-1)], 1, sd, na.rm=TRUE) /
          rowSums(!is.na(acropetal[,2:(ncol(acropetal)-1)]))  )
acropetal$nSEM = acropetal$mean - 
      ( apply( acropetal[,2:(ncol(acropetal)-2)], 1, sd, na.rm=TRUE) /
          rowSums(!is.na(acropetal[,2:(ncol(acropetal)-2)]))  )
#
basipetal.plot = basipetal[,which(names(basipetal)%in%c("s","mean","pSEM","nSEM")) ] %>% 
  gather(key = "statistic", value = "pixel.int",2:4)
basipetal.plot$Position = "basipetal"
acropetal.plot = acropetal[,which(names(acropetal)%in%c("s","mean","pSEM","nSEM")) ] %>% 
  gather(key = "statistic", value = "pixel.int",2:4)
acropetal.plot$Position = "acropetal"
trace.plot = rbind(basipetal.plot, acropetal.plot)

# Plot mean and SEM
ggplot(trace.plot, aes(x = s, y=pixel.int, color=Position, lineend=statistic ))+
    geom_line() +
    #geom_line(aes(y = pSEM),alpha=0.5,color="blue") +
    #geom_line(aes(y = nSEM),alpha=0.5,color="blue") +
    ylab ("Pixel intensity [A.U.]") + 
    xlab ("Time (s)") + 
    ggtitle( paste("GCaMP6 fluorescence. n=",ncol(basipetal)-4," mean±SEM", sep = "") ) +
    theme(
      axis.text.x = element_text(color = "black", size = 20, angle = 0, hjust = 0.5, vjust = 0, face = "plain"),
      axis.text.y = element_text(color = "black", size = 20, angle = 0, hjust = 0 , vjust = 0.5, face = "plain"),
      axis.title.x = element_text(color = "black", size = 25, angle = 0, hjust = 0.5, vjust = 0, face = "plain"), #element_blank(), #element_text(color = "black", size = 30, angle = 0, hjust = 0.5, vjust = 0, face = "plain"), #
      axis.title.y = element_text(color = "black", size = 25, angle = 90, hjust = 0.5, vjust = 1, face = "plain"),
      plot.title = element_text(color = "black", size = 30, angle = 0, hjust = 0.5, vjust = 0, face = "plain"), # element_blank(), 
      legend.text = element_text(size = 20),
      legend.title = element_text(hjust = 0.1, size = 20),
      panel.grid.major.y = element_line(), # element_blank()
      panel.grid.minor.y = element_line(),
      panel.grid.major.x = element_line(), #element_line()
      panel.grid.minor.x = element_blank()) 
ggsave("Pixel.intensity-Mean±SEM.pdf")

```

